---
title: "Static App"
author: "Bryan Sandor"
format: html
editor_options: 
  chunk_output_type: console
---

# Preamble
For this section we will write "static" code that will be used to build an interactive `shiny` app. I have selected data highlighting the scores of the USA National Football League (NFL) for the years 2010 - 2025. The source for this data is found [here](https://www.kaggle.com/datasets/keonim/nfl-game-scores-dataset-2017-2023). We begin by loading the necessary libraries.

```{r load_libraries}
library(tidyverse)
```


# Data Import and Cleaning
First we import the data, removing all rows containing missing values, and then preview it:

```{r data_import}
football <- read_csv("data/2010-2025_scores.csv")
football <- drop_na(football)

football
```

We create one more column indicating the victor in a given game.

```{r game_winner}
football <- add_column(football, "Winner" = "")

for (i in 1:length(football$Winner)){
  if (football$AwayScore[i] > football$HomeScore[i]){
    football$Winner[i] = football$AwayTeam[i]
  } else if (football$AwayScore[i] == football$HomeScore[i]){
    football$Winner[i] = "TIE"
  } else {
    football$Winner[i] = football$HomeTeam[i]
  }
}
```

We will then add a `Date` column, combining the `GameDate` "prefix" and the `Season` "suffix" to the appropriate `date` type format. Note there are several entries with an `NA` designation and thus won't correctly parse.

```{r date_convert}
football <- football |>
              mutate(GameDate = paste(GameDate, Season))

football$GameDate <- mdy(football$GameDate)
```

We then need to adjust some of the `chr` columns to factors.

```{r create_categorical_variables}
football$Season <- as.factor(football$Season)
football$Week <- as.factor(football$Week)
football$GameStatus <- as.factor(football$GameStatus)
football$GameSlot <- as.factor(football$GameSlot)
football$AwayTeam <- as.factor(football$AwayTeam)
football$HomeTeam <- as.factor(football$HomeTeam)
football$Winner <- as.factor(football$Winner)

football
```

# Exploratory Data Analysis
## Victory Counts

First, let's see a count of the total wins for this set of data, sorted in descending order.

```{r win_counts}
football |> group_by(Winner) |>
  summarize(count = n()) |>
  arrange(desc(count)) |>
  print(n = Inf)
```

Tom Brady played with New England for ten years of this data set, viz. 2010 - 2019. I was never a fan, but the guy sure can football a football. Alas, poor Cardinals...

There is a significant advantage for the Home Team, so now let us examine how many times the home team was the victor across the data set.

```{r home_field_advantage}
#| class-output: "wide-output" # fits results in scrollable sub-window

football |> select(HomeTeam, Winner) |>
  table() |>
  print(n = Inf)
```

For an example, Green Bay won 122 of the games for which it was the home team, out of a total number of 207 victories, yielding a proportion of $\frac{122}{207} \approx 0.589$. Note in a binomial experiment with probability of success $p = 0.5$, the probability of at least 122 victories in 207 games is $P(X \geq 122) =$`pbinom(q = 122, size = 207, p = 0.5, lower.tail = FALSE)`$\approx 0.0041$, suggesting a significant home field advantage.