---
title: "Static App"
author: "Bryan Sandor"
format: html
editor_options: 
  chunk_output_type: console
---

# Preamble
For this section we will write "static" code that will be used to build an interactive `shiny` app. I have selected data highlighting the scores of the USA National Football League (NFL) for the years 2010 - 2025. The source for this data is found [here](https://www.kaggle.com/datasets/keonim/nfl-game-scores-dataset-2017-2023). We begin by loading the necessary libraries.

```{r load_libraries}
library(tidyverse)
library(ggplot2)
```


# Data Import and Cleaning
First we import the data, remove all rows containing missing values (which makes the `GameStatus` column redundant, as all scores were listed as `FINAL` or `NA`, so we will delete it). We will be analyzing scores by teams, so records where the home or away team are listed as `TBD` will also be expunged.

```{r data_import}
football <- read_csv("footballApp/data/2010-2025_scores.csv",
                     show_col_types = FALSE)
football <- drop_na(football)
football <- football |> select(-GameStatus)
football <- football |> filter(HomeTeam != "TBD" & AwayTeam != "TBD")

football
```

Within the data, Jacksonville was abbreviated `JAC` until 2013, when it was voted to change their abbreviation to the current `JAX`. Since this is the same team, we will change all the `JAC` abbreviations to `JAX`.

```{r relabel_JAC}
for (i in 1:length(football$HomeTeam)){
  if (football$HomeTeam[i] == "JAC"){
    football$HomeTeam[i] <- "JAX"
  }
  
  if (football$AwayTeam[i] == "JAC"){
    football$AwayTeam[i] <- "JAX"
  }
}
```

We create one more column indicating the victor in a given game. Ties will be calculated as `NA` and will be removed.

```{r game_winner}
football <- add_column(football, "Winner" = "")

for (i in 1:length(football$Winner)){
  if (football$AwayScore[i] > football$HomeScore[i]){
    football$Winner[i] = football$AwayTeam[i]
  } else if (football$AwayScore[i] == football$HomeScore[i]){
    football$Winner[i] = NA
  } else {
    football$Winner[i] = football$HomeTeam[i]
  }
}

football <- drop_na(football)
```

We will then add a `Date` column, combining the `GameDate` "prefix" and the `Season` "suffix" to the appropriate `date` type format. Note there are several entries with an `NA` designation and thus won't correctly parse.

```{r date_convert}
football <- football |>
              mutate(GameDate = paste(GameDate, Season))

football$GameDate <- mdy(football$GameDate)
```

We then need to adjust some of the `chr` columns to factors.

```{r create_categorical_variables}
football$Season <- as.factor(football$Season)
football$Week <- as.factor(football$Week)
football$GameSlot <- as.factor(football$GameSlot)
football$AwayTeam <- as.factor(football$AwayTeam)
football$HomeTeam <- as.factor(football$HomeTeam)
football$Winner <- as.factor(football$Winner)

football
```

# Exploratory Data Analysis
## Numeric
### Qualitative Analyses: Victory Counts

First, let's see a count of the total wins for this set of data, sorted in descending order.

```{r win_counts}
winner_counts <- football |> group_by(Winner) |>
                   summarize(count = n())

print(winner_counts, n = Inf)
```

Tom Brady played with New England for ten years of this data set, viz. 2010 - 2019. I was never a fan, but the guy sure can football a football. Alas, poor Cardinals...

Next, let us examine a count of the number of times a team was the home team.

```{r home_team_counts}
home_counts <- football |> group_by(HomeTeam) |>
                 summarize(count = n())

print(home_counts, n = Inf)
```

There has been shown to be a significant advantage for the Home Team, so now let us examine how many times the home team was the victor across the data set.

```{r home_field_advantage}
football |> select(HomeTeam, Winner) |>
  table() |>
  addmargins() |>
  print(n = Inf)
```

For an example, Green Bay won $W = 122$ of the games for which it was *also* the home team, out of a total number of $H = 166$ home games, yielding a proportion of $P(W | H) = \frac{122}{166} \approx 0.7349$. Note in a binomial experiment with probability of success $p = 0.5$, the probability of at least $W = 122$ victories in $n = 166$ games is $P(X \geq 122) =$`pbinom(q = 122, size = 166, p = 0.5, lower.tail = FALSE)`$\approx$`1.9916e-10`, suggesting a significant home field advantage.

For later, let's also examine, by team, the number of wins they have had as the home team. (This is just the "diagonals" of the previous contingency table.)

```{r home_win_counts}
home_winner_counts <- football |>
  group_by(HomeTeam)|>
  summarize(count = sum(HomeTeam == Winner))
```

And finally, let's compute the total number of games played by each team across all the seasons.

```{r total_counts}
away_counts <- football |> group_by(AwayTeam) |>
  summarize(count = n())

total_game_counts <- data.frame(Team = away_counts[[1]],
                                count = away_counts[[2]] +
                                  home_counts[[2]])

rm(away_counts)
```


### Quantitative Analyses: Scores
Let's explore the home and away scores, overall.

```{r score_summaries}
summary(football$HomeScore)
summary(football$AwayScore)
```

We could, of course, also group by each team and look at each one's overall score summary irrespective of home or away status.

```{r score_summaries_by_team}
football |> group_by("Team" = Winner) |>
  summarize(
    min = min(c(HomeScore, AwayScore)),
    quartile1 = quantile(c(HomeScore, AwayScore), 0.25),
    median = median(c(HomeScore, AwayScore)),
    quartile3 = quantile(c(HomeScore, AwayScore), 0.75),
    max = max(c(HomeScore, AwayScore)),
    IQR = quartile3 - quartile1,
    mean = mean(c(HomeScore, AwayScore)),
    sd = sd(c(HomeScore, AwayScore))
  ) |>
  print(n = Inf)
```

For the most part, many of the medians are close to the means, suggesting symmetric distributions. It is also interesting to note the similarity among the teams themselves, which ostensibly suggests across these seasons the teams are equally matched.

## Graphical
1. Let's first compare the home scores against the away scores.

```{r home_vs_away_scores}
ggplot(data = football, mapping = aes(x = HomeScore, y = AwayScore)) +
  geom_point(position = "jitter") +
  labs(x = "Home Score", y = "Away Score",
       title = "Home vs. Away Scores for NFL",
       subtitle = "(2010 - 2025)")
```

There does appear to be a uniform distribution among these scores.

2. Let's now examine the frequency of the games by day of the week.

```{r game_day_of_week}
ggplot(data = football, mapping = aes(x = GameSlot)) +
  geom_bar() +
  labs(x = "Day/Time Slot", y = "Count",
       title = "Number of NFL Games by Day/Time",
       subtitle = "(2010 - 2025)")
```

Not surprisingly, Sunday dominates, followed (distantly) by Saturday and Thursday, respectively. I was not aware so few Tuesday games existed.

3. Let us now focus on looking at individual teams' statistics.

First we make a data frame recording each team, the number of home games, wins, wins at home, and total games played.

```{r stats_by_team}
team_win_home <- data.frame(Team = as.character(home_counts[[1]]),
                            Wins = winner_counts[[2]],
                            HomeGames = home_counts[[2]],
                            HomeWinGames = home_winner_counts[[2]],
                            TotalGames = total_game_counts[[2]],
                            row.names = NULL)

team_win_home <- as_tibble(team_win_home)
```

Let's explore this home field advantage using a bar graph, with a horizontal line marking the $50-50$ cutoff.

```{r win_given_home_barplot}
ggplot(data = team_win_home,
       mapping = aes(x = Team,
                     y = HomeWinGames/HomeGames)) +
  geom_col(position = "stack", stat = "identity") +
  labs(x = "Team", y = "probability of winning given at home",
       title = "Home Field Win Analysis") +
  geom_abline(aes(intercept = 0.5, slope = 0))
```

The author admits this result surprises him, as he had expected all of the bars to exceed $0.5$.